---
title: "ANA 515 Assignment 3 Getting and Cleaning Data"
author: "Abdul Moid"
date: "2023-06-24"
output: html_document
---

```{r code, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)

```{r code1, echo=TRUE}

#Load Library
library(tidyverse)

# 1. Import CSV File

#Below command used to Import and read the csv data which i saved on my local in my documents

my_data <- read_csv("C:/Users/mfhus/OneDrive/Documents/StormEvents_details-ftp_v1.0_d2016_c20220719.csv")

# 2. Limit the dataframe to the following columns

library(dplyr)

selected_columns <- my_data %>%
  select(BEGIN_YEARMONTH, EPISODE_ID, STATE, STATE_FIPS, CZ_NAME, CZ_TYPE, CZ_FIPS, EVENT_TYPE)

print(selected_columns)


```{r code2, echo=TRUE}

# 3. Arrange the data by the state name (STATE) in ascending order

arranged_data <- selected_columns %>%
  arrange(STATE)
print(arranged_data)

```{r code3, echo=TRUE}

# 4. Change state and county names to title case

library(stringr)

formatted_data <- arranged_data %>%
  mutate(STATE = str_to_title(STATE),
         CZ_NAME = str_to_title(CZ_NAME))

#Print the names to title case
print(formatted_data)

```{r code4, echo=TRUE}

#  5.	Limit to the events listed by county FIPS (CZ_TYPE of “C”) and then remove the CZ_TYPE column 

filtered_data <- formatted_data %>%
  filter(CZ_TYPE == "C") %>%
  select(-CZ_TYPE)

print(filtered_data)

```{r code5, echo=TRUE}

# 6. Pad the state and county FIPS with a “0” at the beginning (hint: there’s a function in stringr to do this) and then unite the two columns to make one FIPS column with the new state-county FIPS code 

library(stringr)
library(tidyr)

padded_data <- filtered_data %>%
  mutate(STATE_FIPS = str_pad(STATE_FIPS, width = 2, side = "left", pad = "0"),
         CZ_FIPS = str_pad(CZ_FIPS, width = 3, side = "left", pad = "0")) %>%
  unite(FIPS, STATE_FIPS, CZ_FIPS, sep = "-")

print(padded_data)

```{r code6, echo=TRUE}

# 7.	Change all the column names to lower case (you may want to try the rename_all function for this) 

lowercase_data <- padded_data %>%
  rename_all(tolower)

print(lowercase_data)

```{r code7, echo=TRUE}

# 8. There is data that comes with base R on U.S. states (data("state")). Use that to create a dataframe with these three columns: state name, area, and region 

# Load the state data
data("state")

# Create a dataframe with state name, area, and region
state_df <- data.frame(
  state_name = state.name,
  area = state.area,
  region = state.region
)

# Print the resulting dataframe
print(state_df)


```{r code8, echo=TRUE}

# 9.	Create a dataframe with the number of events per state in the year of your birth. Merge in the state information dataframe you just created in step 8. Remove any states that are not in the state information dataframe. 

# Load the required packages
library(dplyr)
library(readr)

# Read the storm events dataset
storm_data <- read_csv("C:/Users/mfhus/OneDrive/Documents/StormEvents_details-ftp_v1.0_d2016_c20220719.csv")

# Extract the year from the BEGIN_DATE_TIME column
storm_data$year <- format(as.Date(storm_data$BEGIN_DATE_TIME, "%m/%d/%Y %H:%M:%S"), "%Y")

# Specify the year you're interested in
birth_year <- "1990"

# Filter the dataset for the specific year
storm_data_year <- filter(storm_data, year == birth_year)

# Count the number of events per state
events_per_state <- storm_data_year %>%
  group_by(STATE) %>%
  summarise(Event_Count = n())

# Read the state information dataframe
state_info <- read_csv("path_to_state_information.csv")  # Replace with the actual path to the state information file

# Merge the event count with the state information dataframe
merged_data <- merge(events_per_state, state_info, by = "STATE", all.x = TRUE)

# Remove any states not present in the state information dataframe
merged_data <- na.omit(merged_data)

# Print the final dataframe
print(merged_data)

```{r code9, echo=TRUE}

# 10. Create the following plot 

# Load the required packages
library(ggplot2)
library(readr)

# Create a plot of BEGIN_YEARMONTH, MAGNITUDE, and STATE
plot <- ggplot(storm_data, aes(x = BEGIN_YEARMONTH, y = MAGNITUDE, color = STATE)) +
  geom_point() +
  labs(title = "Storm Events by Year-Month, Magnitude, and State",
       x = "Year-Month",
       y = "Magnitude",
       color = "State") +
  theme(legend.position = "bottom")

# Display the plot
print(plot)





